---
title: "Guide to study design"
description: >
  How to design an animal tracking project with `movedesign`
  (for population-level inferences)
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    number-depth: 3
    theme: cosmo
    html-math-method: katex
    css: assets/custom.scss
    include-in-header:
      text: |
        <link rel="preconnect" href="https://fonts.googleapis.com">
        <link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
        <script>
        // Scroll-reveal for annotated blocks and cards
        document.addEventListener("DOMContentLoaded", () => {
          const els = document.querySelectorAll(
            ".annotation, .design-card, .pipeline-vertical"
          );
          const obs = new IntersectionObserver(
            (entries) => entries.forEach(e => {
              if (e.isIntersecting) {
                e.target.classList.add("visible");
                obs.unobserve(e.target);
              }
            }),
            { threshold: 0.12 }
          );
          els.forEach(el => {
            el.classList.add("reveal");
            obs.observe(el);
          });
        });
        </script>
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

## Overview

<p class="quotes">

Planning an animal tracking study often comes down to three practical
questions: <em>[how many animals should we tag]{style="color:var(--sea);"},
[for how long]{style="color:var(--sea);"}, and
[how frequently]{style="color:var(--sea);"}?</em> Poor design choices
can lead to highly uncertain estimates of key movement metrics such as
<em>home range area</em>, <em>movement speed</em>, and
<em>distance travelled</em>, limiting the reliability of ecological
inferences.

</p>

`movedesign` is available as both an `R` package and a `Shiny`
application and supports the planning and evaluation of movement ecology
studies. Under the hood, `movedesign` simulates tracking datasets using
movement models fitted either to existing tracking data or to
user-defined movement parameters. For each candidate study design it
evaluates how accurately the simulated data recover the true quantity of
interest. This framework lets users examine the consequences of
different design choices — including the number of tagged individuals
([*population* sample size]{style="color:var(--sea);"}), study duration
([*sampling duration*]{style="color:var(--sea);"}), and the time between
successive relocations
([*sampling interval*]{style="color:var(--sea);"}).

This tutorial demonstrates a complete workflow for designing a study to
estimate the mean **home range area** of a tracked population of African
buffalo (*Syncerus caffer*) using empirical tracking data. The examples
are configured to run quickly for demonstration; in practice your
workflows will be more computationally intensive.

```{=html}
<div class="annotation">
  <span class="annotation-icon">💬</span>
  <div>At its core, this tutorial addresses one question: given what we
  know about how this species moves, what study design — how many
  individuals, tracked for how long, and at what sampling interval —
  lets us reliably estimate mean home range area?</div>
</div>
```

The workflow proceeds in six steps:

```{=html}
<div class="pipeline-vertical">

  <div class="pipeline-v-step">
    <div class="pipeline-v-num">1</div>
    <div class="pipeline-v-content">
      <a href="#step-1" class="pipeline-v-fn">data() / fitting_models()</a>
      <span class="pipeline-v-label">Load tracking data and fit movement models</span>
    </div>
  </div>

  <div class="pipeline-v-step">
    <div class="pipeline-v-num">2</div>
    <div class="pipeline-v-content">
      <a href="#step-2" class="pipeline-v-fn">md_prepare()</a>
      <span class="pipeline-v-label">Prepare simulation inputs for competing designs</span>
    </div>
  </div>

  <div class="pipeline-v-step">
    <div class="pipeline-v-num">3</div>
    <div class="pipeline-v-content">
      <a href="#step-3" class="pipeline-v-fn">md_run()</a>
      <span class="pipeline-v-label">Run simulations and evaluate design performance</span>
    </div>
  </div>

  <div class="pipeline-v-step">
    <div class="pipeline-v-num">4</div>
    <div class="pipeline-v-content">
      <a href="#step-4" class="pipeline-v-fn">md_plot_preview() / md_compare_preview()</a>
      <span class="pipeline-v-label">Visualise and compare candidate designs</span>
    </div>
  </div>

  <div class="pipeline-v-step">
    <div class="pipeline-v-num">5</div>
    <div class="pipeline-v-content">
      <a href="#step-5" class="pipeline-v-fn">md_replicate()</a>
      <span class="pipeline-v-label">Run multiple replicates for robust outputs</span>
    </div>
  </div>

  <div class="pipeline-v-step">
    <div class="pipeline-v-num">6</div>
    <div class="pipeline-v-content">
      <a href="#step-6" class="pipeline-v-fn">md_check() / md_compare()</a>
      <span class="pipeline-v-label">Diagnose and report final results</span>
    </div>
  </div>

</div>
```

### Background

`movedesign` builds on **continuous-time movement models** fitted via
the `ctmm` package. These models describe animal movement as a
stochastic process governed by two key timescales:

- **τ**~p~ — [position autocorrelation
  timescale]{style="color:var(--sea);"}: how long it takes an animal to
  cross the linear extent of its home range (the *average home range
  crossing time*). This timescale is most directly relevant to this
  workflow — your sampling duration must capture enough home-range
  crossing events to reliably estimate home range area.

- **τ**~v~ — [velocity autocorrelation
  timescale]{style="color:var(--sea);"}: how long speed and direction
  remain consistent (the *average directional persistence*). A short
  τ~v~ describes erratic, undirected movement; a long τ~v~ describes an
  animal moving in sustained, directed bouts.

Both are estimated from empirical data and used to simulate realistic
movement data for study-design evaluation.

### Data

The built-in `buffalo` dataset (available within `ctmm`) contains GPS
tracks for **six African buffalo** (*Syncerus caffer*) from Kruger
National Park, South Africa. This dataset is a named list of `telemetry`
objects, one per individual.

## Workflow

```{r}
#| label: load-models
#| echo: false
#| results: hide

start_time <- Sys.time()

library(movedesign)
library(ctmm)
library(here)
library(dplyr)

data(buffalo)

models_init <- readRDS(
  here::here("..", "data", "processed", "buffalo_fitList.rds"))
models <- list()
for (i in seq_along(models_init)) {
  models[[i]] <- models_init[[i]][[1]]
}

```

### Load packages, data and fit models {#step-1}

```{r}
#| label: install
#| eval: false

# Install packages from CRAN:
install.packages(c("ctmm", "movedesign"))
```

```{=html}
<div class="annotation">
  <span class="annotation-icon">💬</span>
  <div>Load <code>movedesign</code> for the study-design workflow and
  <code>ctmm</code> for the movement models and tracking data. Both must
  be installed from CRAN before running the rest of this tutorial.</div>
</div>
```

```{r}
#| label: setup
#| eval: false

# Load packages:
library(movedesign)
library(ctmm)

# Load the buffalo dataset bundled with ctmm:
data(buffalo)

# Fit movement models (may take several minutes):
models <- fitting_models(buffalo, .parallel = FALSE)
```

`fitting_models()` performs model selection across candidate movement
models — Brownian motion, Ornstein-Uhlenbeck, and their variants — and
identifies the best fit for each individual. Save the output with
`saveRDS()` so you can reload it without repeating the computation.

```{r}
#| label: save
#| eval: false

saveRDS(models, "models_buffalo.rds")
```

When complete, `models` will be a list of lists: one element per
individual, each containing that individual's best-fit movement model.

### Prepare simulation inputs {#step-2}

`md_prepare()` builds the simulation blueprint for a given study design.
It validates inputs, checks that movement models are available (fitting
them automatically if needed), and extracts the relevant movement
parameters (τ~p~, τ~v~). These are combined with your design choices:
number of individuals, sampling duration, sampling interval, and
grouping structure.

```{=html}
<div class="annotation">
  <span class="annotation-icon">💬</span>
  <div>The table below describes every argument in
  <code>md_prepare()</code>. Run <code>?md_prepare()</code> for the
  full documentation.</div>
</div>
```

| Argument | Type | Description |
|---|---|---|
| `species` | character | Label for the species; appears in plots and summaries. |
| `data` | list | Tracking data (list of `telemetry` objects). |
| `models` | list | Fitted models, one per individual. |
| `n_individuals` | integer | Population sample size. |
| `dur` | named list | Sampling duration: `list(value = 1, unit = "month")`. |
| `dti` | named list | Sampling interval: `list(value = 5, unit = "days")`. |
| `add_individual_variation` | logical | Draw parameters from the population distribution. |
| `set_target` | character | `"hr"` = home range; `"ctsd"` = speed & distance. |
| `which_meta` | character | `"mean"` = population mean; `"ratio"` = group ratio. |
| `groups` | named list | *(Grouped designs only)* assigns individuals to group labels. |
| `parallel` | logical | Enable parallel processing (strongly recommended). |

: {tbl-colwidths="[20,20,60]"}

We compare two designs, each estimating the *population mean* **home
range area** under different sampling parameters:

```{=html}
<div class="design-grid">

  <div class="design-card">
    <div class="design-card-header design-a">
      Design A
      <span class="design-card-badge">Short · Fine</span>
    </div>
    <div class="design-card-body">
      <div class="design-stat">
        <span class="design-stat-label">Sampling duration</span>
        <span class="design-stat-value">1 month</span>
      </div>
      <div class="design-stat">
        <span class="design-stat-label">Sampling interval</span>
        <span class="design-stat-value">every 5 days</span>
      </div>
      <div class="design-stat">
        <span class="design-stat-label">No. of individuals</span>
        <span class="design-stat-value">10</span>
      </div>
      <p class="design-tradeoff tradeoff-a">Short duration but finer
      resolution, with a moderate population sample size
      (<em>m</em>).</p>
    </div>
  </div>

  <div class="design-card">
    <div class="design-card-header design-b">
      Design B
      <span class="design-card-badge">Long · Coarse</span>
    </div>
    <div class="design-card-body">
      <div class="design-stat">
        <span class="design-stat-label">Sampling duration</span>
        <span class="design-stat-value">2 years</span>
      </div>
      <div class="design-stat">
        <span class="design-stat-label">Sampling interval</span>
        <span class="design-stat-value">every 10 days</span>
      </div>
      <div class="design-stat">
        <span class="design-stat-label">No. of individuals</span>
        <span class="design-stat-value">6</span>
      </div>
      <p class="design-tradeoff tradeoff-b">Longer duration and coarser
      resolution, with a smaller population sample size
      (<em>m</em>).</p>
    </div>
  </div>

</div>
```

#### Design A

```{r}
#| label: prepare-hrA
#| results: hide
#| message: false
#| warning: false

input_hrA <- md_prepare(
  species                  = "buffalo",
  data                     = buffalo,
  models                   = models,
  dur                      = list(value = 1, unit = "month"),
  dti                      = list(value = 5, unit = "days"),
  n_individuals            = 10,
  add_individual_variation = TRUE,
  set_target               = "hr",
  which_meta               = "mean",
  parallel                 = TRUE)
```

#### Design B

```{r}
#| label: prepare-hrB
#| results: hide
#| message: false
#| warning: false

input_hrB <- md_prepare(
  species                  = "buffalo",
  data                     = buffalo,
  models                   = models,
  dur                      = list(value = 2,  unit = "years"),
  dti                      = list(value = 10, unit = "days"),
  n_individuals            = 6,
  add_individual_variation = TRUE,
  set_target               = "hr",
  which_meta               = "mean",
  parallel                 = TRUE)
```

Inspect the prepared objects before running simulations:

```{r}
#| label: inspect-inputsA
#| eval: false

summary(input_hrA)
summary(input_hrB)
```

```{r}
#| label: print-inputA
#| echo: false

print(input_hrA)
```

```{r}
#| label: print-inputB
#| echo: false

print(input_hrB)
```

### Run simulations {#step-3}

`md_run()` runs one full instance of the study design workflow and
evaluates how well it recovers the target quantity. Because a single run
is subject to stochastic variation (especially when
`add_individual_variation = TRUE`), treat its outputs as exploratory.
Later, `md_replicate()` produces robust results for reporting.

```{r}
#| label: run-sims
#| results: hide
#| message: false
#| warning: false

output_run_hrA <- md_run(input_hrA)
output_run_hrB <- md_run(input_hrB)
```

```{r}
#| label: inspect-outputs
#| eval: false

summary(output_run_hrA)
summary(output_run_hrB)
```

```{r}
#| label: print-outputsA
#| echo: false

print(output_run_hrA)
```

```{r}
#| label: print-outputsB
#| echo: false

print(output_run_hrB)
```

### Preview design performance {#step-4}

`md_plot_preview()` shows how estimation error (relative error in home
range area) changes with population sample size. It provides preliminary
insight into design performance from a single stochastic run.

By default the plot shows estimates for a random combination of
individuals at each sample size. Small sample sizes can produce highly
unstable results.

```{r}
#| label: preview-hrB-default
#| warning: false
#| message: false
#| fig.height: 4
#| out.width: "100%"

md_plot_preview(output_run_hrB)
```

If `n_resamples` is specified, the function evaluates up to `n` unique
combinations at each population sample size, summarising results as mean
relative error with confidence intervals. In design B with
`n_individuals = 6`, the smallest sample size (selecting 2 from 6)
yields $\binom{6}{2} = 15$ possible combinations.

```{r}
#| label: preview-resampled
#| warning: false
#| message: false
#| fig.height: 4
#| out.width: "100%"

# Explicit resampling:
md_plot_preview(output_run_hrB, n_resamples = 15)
```

Variability narrows at larger sample sizes because the number of unique
combinations shrinks as `k` approaches `n`.

#### Compare candidate designs

`md_compare_preview()` plots estimation error side by side for two or
more designs. The `error_threshold` argument adds a reference line
marking the maximum acceptable relative error — forcing an explicit
decision about how much estimation error is tolerable. Here we set it to
`0.1` (10%).

```{r}
#| label: compare-designs
#| eval: false

md_compare_preview(
  list(output_run_hrA, output_run_hrB),
  error_threshold = 0.1,
  n_resamples     = 15)
```

```{r}
#| label: compare-designs-seeded
#| echo: false
#| results: hide
#| fig.height: 6
#| out.width: "100%"

out_preview <- md_compare_preview(
  list(output_run_hrA, output_run_hrB),
  error_threshold = 0.1,
  n_resamples     = 15)

err_A <- dplyr::pull(
  dplyr::slice_max(out_preview[[1]], m), error)
err_B <- dplyr::pull(
  dplyr::slice_max(out_preview[[2]], m), error)

dir_A <- ifelse(err_A < 0, "underestimated", "overestimated")
dir_B <- ifelse(err_B < 0, "underestimated", "overestimated")

err_A <- round(err_A * 100, 0)
err_B <- round(err_B * 100, 0)
```

In this example, design A `r dir_A` home range area by around
`r abs(err_A)`% at `r output_run_hrA$n_individuals` individuals
(`r ifelse(abs(err_A) > 15, "outside", "within")` the threshold), while
design B `r dir_B` by around `r abs(err_B)`% with
`r output_run_hrB$n_individuals` individuals
(`r ifelse(abs(err_B) > 15, "outside", "within")` the threshold).

Your results may differ — variability between runs is expected and is
precisely why `md_run()` outputs should be treated as preliminary.

### Run multiple replicates for robust outputs {#step-5}

`md_replicate()` runs the workflow many times, each time drawing an
independent set of simulated individuals under the specified design.
Think of each replicate as a hypothetical version of your study.
Aggregating across many such hypothetical studies reveals what you can
reliably expect: how close estimates tend to be to the truth, and how
much they vary.

```{=html}
<div class="annotation">
  <span class="annotation-icon">💬</span>
  <div><code>n_replicates = 5</code> is used here for illustration — it
  is the minimum recommended number. In practice, reliable results
  typically require considerably more replicates.</div>
</div>
```

```{r}
#| label: replicate
#| results: hide
#| message: false
#| warning: false

out_replicate_hrA <- md_replicate(input_hrA, n_replicates = 5)
out_replicate_hrB <- md_replicate(input_hrB, n_replicates = 5)
```

Once complete, inspect each design using `md_plot()` and `md_check()`:

```{r}
#| label: plot-replicates
#| eval: false

md_plot(out_replicate_hrA)
md_plot(out_replicate_hrB)
```

```{r}
#| label: check-replicates
#| eval: false

md_check(out_replicate_hrA, n_converge = 4)
md_check(out_replicate_hrB, n_converge = 4)
```

### Report {#step-6}

`md_compare()` takes the replicated outputs from two or more designs and
produces a density plot of relative error across replicates, giving a
full picture of performance. Designs are ranked by combining the error
distribution with the width and position of the credible interval
relative to zero. The best-performing design is identified automatically.

```{r}
#| label: compare
#| fig.height: 8
#| out.width: "100%"
#| message: false
#| warning: false

out_report <- md_compare(list(out_replicate_hrA,
                              out_replicate_hrB))
```

```{r}
#| label: compare-seeded
#| echo: false
#| results: hide
#| fig.height: 6
#| out.width: "100%"

ranking      <- out_report$ranking
joint_winners <- out_report$winners

winner <- joint_winners[
  joint_winners$type == "hr", , drop = FALSE]$design_id

rank_rows <- ranking[
  ranking$design_id == winner &
    ranking$type == "hr" &
    ranking$rank == 1L, , drop = FALSE]
```

```{r}
#| label: inspect-report
#| eval: false

summary(out_report, verbose = TRUE)
```

```{r}
#| label: print-report
#| echo: false

print(out_report, verbose = TRUE)
```

In this workflow, the best design was `r winner`, yielding a relative
error of `r round(rank_rows$error * 100, 1)`% \[95% CI:
`r round(rank_rows$error_lci * 100, 1)`,
`r round(rank_rows$error_uci * 100, 1)`%\], with `r rank_rows$m` tagged
individuals tracked for `r rank_rows$dur` `r rank_rows$dur_unit` every
`r rank_rows$dti` `r rank_rows$dti_unit`.

## Session Info

```{r}
#| echo: false

runtime <- Sys.time() - start_time
cat("Total run time:", format(runtime), "\n")
```

```{r}
#| label: session-info

utils::sessionInfo()
```
